Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2015-01-30 draft_lm1.tex
Small fix in eqs 4/5
another review of the text

Dengue transmission is modulated by environmental conditions, of which 
temperature, due to its effects on the 
vector reproduction, stands out as a strong predictor of 
incidence~\cite{honorio_temporal_2009,wu_higher_2009}.
In places with sufficient temperature variation dengue is predominantly a 
summer disease. So it is fair to say that these environmental fluctuations play 
a key role in determining beginning and end of epidemic periods through its 
effects on the force of infection, which cannot be taken as 
constant\cite{reiner_time-varying_2014}

@@

The Attack ratio of disease is a measure of morbidity defined as the number of 
new cases divided by the population at risk.
For Dengue outbreaks and epidemics this can be difficult to calculated due to 
the lack of information about the population at risk. The population at risk in 
the case of dengue, is the number of susceptibles to the circulating virus 
type(s) before a given epidemic.

The main source of information about Dengue dynamics are the incidence 
time series, which represent all the cases of Dengue reported by health 
professionals with no discrimination of virus types.
Only a very small sample 
of the cases are ever examined to determined the virus type.

@@

The attack ratio is also influenced by the reproductive number of the disease 
~\cite{bacaer_final_2009, katriel_attack_2012}.
With Dengue, since transmission 
is affected by environmental conditions as discussed above, the reproductive 
number is also expected to vary with time.

Various challenges exist to assess the attack rate of a dengue epidemic. 
One of 
them is that there are a large number of asymptomatic cases in any epidemic, 
which nevertheless acquire immunity.
It is estimated that for 
every case reported, 10-20 are not seen by health authorities (personnal 
communication).
Thirdly, under-reporting is a serious issue in Brazil. 
Duarte and Fran\c{c}a (2006)~\cite{duarte_data_2006}, estimated 
the sensitivity of dengue reporting for hospitalized patients in 
Belo-Horizonte, Brazil to be of 63\%, 
meaning that approximately 37\% of the suspected dengue cases go unreported.  
Lastly, demography and migrations affect the number of susceptible in ways which 
are not easy to determine.

In this paper, we estimate the number of susceptibles before every major 
outbreak ($S_{0, year}$) in the last 18 years.
Due to the alternation of circulating serotypes, we detected fluctuating levels 
of susceptibility the circulating virus(es).
To estimate the susceptibles we have we have fixed the 
transmissibility based on the $R_t$ series derived from data~\cite{nishiura}. 
Then, 
from the incidence series and the population at risk, we could then calculate 
the attack ratio for each epidemic.

To accomplish the estimation of the $S_0$, we use a simplified model of 
dengue transmission based on a single-strain Susceptible-Infecctious-Removed 
(SIR) model.
The reason for the choice of a single strain model, is the lack of 
accurate information regarding the actual proportions of each virus in 
circulation.
However, some information is available about the predominant 
serotypes for some epidemics in the period 
of study~\cite{macedo_virological_2013}.

@@

In this section we will start by describing the data and then the method used 
to estimate the effective reproductive number, $R_t$, from the data and obtain 
its posterior distribution. 
We then proceed to describe Susceptible-infectious-recovered (SIR) model 
used to represent the global disease incidence and how $R_t$ can be integrated 
into the model to allow for time varying transmission. 
Next, an approach to approximate the posterior distributions 
of the numbers of susceptible to the main circulating Dengue viruses for each 
epidemic is detailed.
Finally, we discuss how to estimate the attack ratio of each 
epidemic using the estimated susceptible fraction and observed incidence.

\subsection*{The data set} 
The data used in this paper consists of  Dengue weekly notified cases time 
series for the city of Rio de Janeiro from 1996 to 2014. The cases are notified 
based only on clinical symptoms. Laboratory confirmation and serotype 
information are available only for a very small sample and only on recent 
years (2010-2013). The incidence is normalized for the parameter estimation 
procedures. The normalization was done by dividing the number os cases reported 
by the total city population at each year as given by the census.

@@

We explore the approach of Ederer and Mantel~\cite{mantel}, whose objective is 
to obtain 
confidence intervals for the ratio of two Poisson counts. 
Let $Y_{t} \sim Poisson(\lambda_t)$ and $Y_{t+1} \sim Poisson(\lambda_{t+1})$ 
and define $S = Y_{t} + Y_{t+1}$.
The authors note that by conditioning on the sum $S$
\begin{align}
\label{eq:binlike}
Y_{t+1} | S &\sim Binomial(S, \theta_t) \\
\theta_t &= \frac{\lambda_{t+1}}{\lambda_{t} + \lambda_{t+1}}
\end{align}
Let $c_{\alpha}(\theta_t) = \{\theta_t^{(L)} , \theta_t^{(U)} \}$ be such that 
$Pr(\theta_t^{(L)}<\theta_t <\theta_t^{(U)}) = \alpha$.
Analogously, define $c_{\alpha}(R_t) = \{R_t^{(L)} , R_t^{(U)} \}$ such that 
$Pr(R_t^{(L)}<R_t<R_t^{(U)}) = \alpha$.
Ederer and Mantel (1974)~\cite{mantel} show that one can construct a $100\alpha 
\%$ confidence interval for \rr by noting that
\begin{equation}
\label{eq:confRt}
 R_t^{(L)} = \frac{\theta_t^{(L)}}{(1-\theta_t^{(L)})} \quad \text{and} \quad R_t^{(U)} = \frac{\theta_t^{(U)}}{(1-\theta_t^{(U)})}\nonumber \\
\end{equation}
Because the transform from $\theta$ to \rr is monotonically 
increasing, the result holds for confidence and credibility intervals alike.

@@

Table \ref{tab:AR} contains the attack ratios and medians of the $S_0$ 
estimated for each epidemic/outbreak. It is interesting to notice that the 
larger epidemics, in terms of total number of cases are not the one with the 
greater attack ratios. This stresses the importance of knowing the imunological 
structure of the population in order to truly assess the impact of eny 
epidemic.

>>>>

Dengue transmission is also modulated by environmental conditions, among which, 
temperature, due to its effects on the 
vector reproduction, stands out as a strong predictor of 
incidence~\cite{honorio_temporal_2009,wu_higher_2009}.
In places with sufficient temperature variation dengue is predominantly a 
summer disease. So it is fair to say that these environmental fluctuations play 
a key role in determining beginning and end of epidemic periods through its 
effects on the force of infection, which cannot be taken as 
constant\cite{reiner_time-varying_2014}

@@

The Attack ratio (AR) of a disease is a measure of morbidity defined as the 
number of new cases divided by the population at risk.
For Dengue outbreaks and epidemics, it can be difficult to calculate the AR due 
to the lack of knowledge of the population at risk. The population at risk 
in the case of dengue, is the number of susceptibles to the circulating virus 
type(s) before a given epidemic.

@@

The attack ratio is also influenced by the reproductive number of the disease 
~\cite{bacaer_final_2009, katriel_attack_2012}, which is closely associated 
with the force of infection. Thus the incorporation of the effective 
reproduction number, \rr, as a function of time, is crucial to an accurate 
estimation of the AR.

There are other sources of uncertainty impinging on the estimation of the AR
of a dengue epidemic. One of them is that there are a large number of 
asymptomatic cases in any epidemic, which nevertheless acquire immunity.
It is estimated that for every case reported, 10-20 are not seen by health 
authorities (personal communication).
Another is under-reporting, which is a serious issue in Brazil. 
Duarte and Fran\c{c}a (2006)~\cite{duarte_data_2006}, estimated 
the sensitivity of dengue reporting for hospitalized patients in 
Belo-Horizonte, Brazil to be of 63\%, 
meaning that approximately 37\% of the suspected dengue cases go unreported.  
Lastly, demography and migrations affect the number of susceptible in ways which 
are not easy to fully determine.

In this paper, we estimate the number of susceptibles before every major 
outbreak ($S_{0, year}$) in the last 18 years.
Due to the alternation of circulating serotypes, we detected fluctuating levels 
of susceptibility to the circulating virus(es).
To estimate the susceptibles we have set the 
transmissibility from the $R_t$ series derived from data~\cite{nishiura} 
(\ref{eq:effbeta}). 
Then, 
from the incidence series and the population at risk, we could then calculate 
the attack ratio for each epidemic.

To accomplish the estimation of the $S_0$, we use a simplified model of 
dengue transmission based on a single-strain Susceptible-Infectious-Removed 
(SIR) model.
The reason for the choice of a single strain model, is the lack of 
 information regarding the actual proportions of each virus in 
circulation.
However, some information is available about the predominant 
serotypes for some epidemics in the period 
of study~\cite{macedo_virological_2013}.

@@

In this section we will start by describing the data and then the method used 
to estimate the effective reproductive number, $R_t$, from the data and obtain 
its posterior distribution. 
We then proceed to describe the Susceptible-Infectious-Recovered (SIR) model 
used to represent the aggregated disease incidence and how $R_t$ can be 
integrated into the model to allow for time varying force of infection. 
Next, an approach to approximate the posterior distributions 
of the numbers of susceptible to the main circulating Dengue viruses for each 
epidemic is detailed.
Finally, we discuss how to estimate the attack ratio of each 
epidemic using the estimated susceptible fraction and the observed incidence.

\subsection*{The data set} 
The data used in this paper consists of time 
series of weekly notified cases of Dengue for the 
city of Rio de Janeiro from 1996 to 2014. The cases are notified 
based only on clinical criteria. Laboratory confirmation and serotype 
information are available only for a very small sample and only on recent 
years (2010-2013). The incidence is normalized for the parameter estimation 
procedures. The normalization was done by dividing the number of cases reported 
by the total city population at each year as given by the census.

@@

We explore the approach of Ederer and Mantel~\cite{mantel}, whose objective is 
to obtain 
confidence intervals for the ratio of two Poisson counts. 
Let $Y_{t} \sim Poisson(\lambda_t)$ and $Y_{t+1} \sim Poisson(\lambda_{t+1})$ 
and define $S = Y_{t} + Y_{t+1}$.
The authors note that by conditioning on the sum $S$
\begin{align}
\label{eq:binlike}
Y_{t+1} | S &\sim Binomial(S, \theta_t) \\
\theta_t &= \frac{\lambda_{t+1}}{\lambda_{t} + \lambda_{t+1}}
\end{align}
Let $c_{\alpha}(\theta_t) = \{\theta_t^{(L)} , \theta_t^{(U)} \}$ be such that 
$Pr(\theta_t^{(L)}<\theta_t <\theta_t^{(U)}) = \alpha$.
Analogously, define $c_{\alpha}(R_t) = \{R_t^{(L)} , R_t^{(U)} \}$ such that 
$Pr(R_t^{(L)}<R_t<R_t^{(U)}) = \alpha$.
Ederer and Mantel (1974)~\cite{mantel} show that one can construct a $100\alpha 
\%$ confidence interval for \rr by noting that
\begin{equation}
\label{eq:confRt}
 R_t^{(L)} = \frac{\theta_t^{(L)}}{(1-\theta_t^{(L)})} \quad \text{and} \quad 
R_t^{(U)} = \frac{\theta_t^{(U)}}{(1-\theta_t^{(U)})}\\
\end{equation}
Because the transform from $\theta$ to \rr is monotonically 
increasing, the result holds for confidence and credibility intervals alike.

@@

Table \ref{tab:AR} contains the attack ratios and medians of the $S_0$ 
estimated for each epidemic/outbreak. It is interesting to notice that the 
larger epidemics, in terms of total number of cases are not the one with the 
greater attack ratios. This stresses the importance of knowing the 
immunological structure of the population in order to truly assess the impact 
of any epidemic.
